Relational Parametricity for Algebraically Indexed Types

### 2012-06-17 paper/geometry-twocolumn.tex

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Note how the type of the result of $\mathrm{offset}$ simplifies from
$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$. The algebraic structure on
the indexing theory induces type equalities.  The type of the vector
addition operation can be specialised to the case of moving a point by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}&
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

@@

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors. The standard Euclidean dot
product is defined on the coordinate representation in the following
way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To incorporate the operation of dot product into our calculus, we must
assign it a type. The dot product is not invariant under the action of
either of the groups $\GL{2}$ and $\Transl{2}$ that we have considered
so far. However, the dot product is invariant under the subgroup
$\Orth{2}$ of $\GL{2}$ of \emph{orthogonal} linear transformations:
invertible linear maps whose determinant is either $1$ or $-1$. We
introduce orthogonal transformations as a new sort $\SynOrth{2}$, and
overload the multiplicative group operations syntax for inhabitants of
$\SynOrth{2}$. We also assume a function $\iota_O$ that takes $e :
\SynOrth{2}$ to $\iota_O(e) : \SynGL{2}$. We may now assign a type to
the dot product, using quantification over $\SynOrth{2}$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

>>>>

Given two points that have been constrained to be invariant with
respect to the same change of origin---i.e.~two values of type
$\tyPrim{vec}{t}$---we can compute their offset, which is a vector
expressed with respect to the null change of origin. The offset
operation is a special case of vector subtraction:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{offset} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{t} \to \tyPrim{vec}{0} \\
    \mathrm{offset}\ [t]\ p_1\ p_2 = p_1 - p_2
  \end{array}
\end{displaymath}
Note how the type of the result of $\mathrm{offset}$ simplifies from
$\tyPrim{vec}{t-t}$ to $\tyPrim{vec}{0}$: the algebraic structure on
the indexing theory induces type equalities.  The type of the vector
addition operation can be specialised to the case of moving a point by
a vector:
\begin{displaymath}
  \begin{array}{l}
    \mathrm{moveBy} : \forall t \mathord:\SynTransl{2}.\ \tyPrim{vec}{t} \to \tyPrim{vec}{0} \to \tyPrim{vec}{t} \\
    \mathrm{moveBy}\ [t]\ p\ v = p + v
  \end{array}
\end{displaymath}
The types we assign to the remaining vector space primitives, the zero
vector and multiplication by a scalar, do not describe any interesting
effect on translations:
\begin{displaymath}
  \begin{array}{l@{\hspace{0.5em}:\hspace{0.5em}}l@{\hspace{4em}}l@{\hspace{0.5em}:\hspace{0.5em}}l}
    \mathrm{0} & \tyPrim{vec}{0}&
    (*) & \tyPrimNm{real} \to \tyPrim{vec}{0} \to \tyPrim{vec}{0}
  \end{array}
\end{displaymath}

@@

\paragraph{Euclidean Geometry} Euclidean geometry extends affine
geometry with the operation of \emph{dot product} (also known as the
\emph{inner product}) of two vectors. The standard Euclidean dot
product is defined on the coordinate representation in the following
way:
\begin{displaymath}
  (x_1,y_1) \cdot (x_2,y_2) = x_1x_2 + y_1y_2
\end{displaymath}
% The dot product is used to define properties of vectors such as their
% \emph{norm} $||\vec{v}|| = \vec{v}\cdot\vec{v}$, which is the square
% of the length of the vector $\vec{v}$, and the notion of
% orthogonality: two vectors $\vec{v_1}$ and $\vec{v_2}$ are orthogonal
% if $\vec{v_1}\cdot\vec{v_2} = 0$.
To incorporate the operation of dot product into our calculus, we must
assign it a type. The dot product is not invariant under the action of
either of the groups $\GL{2}$ and $\Transl{2}$ that we have considered
so far. However, the dot product is invariant under the subgroup
$\Orth{2}$ of $\GL{2}$ of \emph{orthogonal} linear transformations:
invertible linear maps whose matrix representation's transpose is
equal to their inverse. We introduce orthogonal transformations as a
new sort $\SynOrth{2}$, and overload the multiplicative group
operations syntax for inhabitants of $\SynOrth{2}$. We also assume a
function $\iota_O$ that takes $e : \SynOrth{2}$ to $\iota_O(e) :
\SynGL{2}$. We may now assign a type to the dot product, using
quantification over $\SynOrth{2}$:
\begin{displaymath}
  (\cdot) : \forall O \mathord: \SynOrth{2}.\ \tyPrim{vec}{\iota_OO, 0} \to \tyPrim{vec}{\iota_OO, 0} \to \tyPrimNm{real}
\end{displaymath}

### 2012-06-17 paper/section-further-examples.tex

>>>>

\subsection{Monoid Indexed Types}
\label{sec:monoid-indexed-types}

Primitive type is a monoid, and we index by a commutative monoid. An
example of this is strings with tainting information, or bags, or sets
of stuff. This gives a slightly different information flow analysis to
the logical information flow above. In particular, we can now count
uses of input data.

Primitive operations $\Gamma_{\mathit{Mon}}$:
\begin{displaymath}
  \begin{array}{c@{\hspace{0.5em}:\hspace{0.5em}}l}
    1 & \tyPrim{M}{1} \\
    (\cdot) & \forall a,b.\ \tyPrim{M}{a} \to \tyPrim{M}{b} \to \tyPrim{M}{a\cdot b}
  \end{array}
\end{displaymath}

For the index-erasure interpretation, we assume a commutative monoid
$\mathcal{M}$. The primitive type $\tyPrimNm{M}$ is interpreted as
$\tySem{\tyPrim{M}{e}} = \mathcal{M}$. The primitive operations in
$\Gamma_{\mathit{Mon}}$ are interpreted just as the unit and multiplication of
the monoid $\mathcal{M}$.

% Need the following definitions: syntatic group/monoid arising from
% the syntax, and a neater way of defining relational
% environments. Also, a neat way of formally introducing a load of
% primitive operations.

For each index context $\relEnv{E}(\Delta)$, the set
$\relEnv{E}(\Delta)$ of relational environments is defined to be $\{
\rho_{(M,h)} \sepbar M\textrm{ is a submonoid of }(\Delta \Rightarrow
\mathsf{M})/\equiv, h\textrm{ is a monoid homomorphism }M \to
Z(\mathcal{M}) \}$, where
\begin{displaymath}
  \rho_{(M,h)}\ \tyPrimNm{M}\ (e) = \left\{
    \begin{array}{ll}
      \{ \} & \textrm{if}\ e\not\in M \\
      \{ (x, h(e)\cdot x) \sepbar x \in \mathcal{M} \} & \textrm{if}\ e\in M
    \end{array}
  \right.
\end{displaymath}
Note the mapping to the centre of the monoid $\mathcal{M}$.

\begin{lemma}
  For all $\Delta$ and $\rho \in \relEnv{E}(\Delta)$,
  $(\eta_{M},\eta_{M}) \in \rsem{\Gamma_{\mathit{Mon}}}{\relEnv{E}}\rho$.
\end{lemma}

\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matricies of
  natural numbers $A$ and $m$-by-$1$ matricies of natural numbers $b$,
  there exists a program $M$ such that
  \begin{equation}
    \label{eq:fo-monoid-type}
    \Gamma_{\mathit{Mon}} \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{M}.\ \tyPrim{M}{e_1} \to \dots \to \tyPrim{M}{e_n} \to \tyPrim{M}{e}
  \end{equation}
  where $e_j = i_1^{A_{j1}}\dots i_m^{A_{jm}}$ and $e =
  i_1^{b_1}\dots i_m^{b_m}$, if and only if there is a solution
  $\vec{x} = (x_1,\cdots,x_n)$ in the natural numbers to the equation
  $A x^\top = b$.
\end{theorem}

\begin{proof}
  (If) The program $\Lambda i_1\dots i_m.\ \lambda z_1\dots z_n.\
  z_1^{x_1}\dots z_n^{x_n}$ satisfies the typing judgment
  (\ref{eq:fo-monoid-type}).

  (Only if) Assume that there is a program $M$ satisfying the typing
  judgment (\ref{eq:fo-monoid-type}). By \thmref{thm:abstraction} we
  know that for all relational environments $\rho \in
  \relEnv{E}(i_1,\dots,i_m)$, and all $z_j,z'_j \in \mathcal{M}$,
  \begin{displaymath}
    \begin{array}{l}
      (z_1,z'_1) \in \rho\ \tyPrimNm{M}\ (e_1) \land \dots \land (z_n,z'_n) \in \rho\ \tyPrimNm{M}\ (e_n) \Rightarrow \\
      \quad (\tmSem{M}z_1\dots z_n, \tmSem{M}z'_1\dots z'_n) \in \rho\ \tyPrimNm{M}\ (e)
    \end{array}
  \end{displaymath}
  We select the relational environments
  \begin{displaymath}
    \rho\ \tyPrimNm{M}\ (e) = \left\{
      \begin{array}{ll}
        \{(z,z) \sepbar z \in \mathcal{M} \} &
        \begin{array}[t]{@{}l}
          \textrm{if}\ e = e_1^{x_1}\dots e_n^{x_n} \\
          \textrm{ for some }x_1,\dots,x_n \in \mathbb{N}
        \end{array}
        \\
        \{\} & \textrm{otherwise}
      \end{array}
    \right.
  \end{displaymath}
  Now, for any $z_1,\dots,z_m \in \mathcal{M}$, we know that, for all
  $j$, $(z_j,z_j) \in \rho\ \tyPrimNm{M}\ (e_j)$, so we know that
  $(\tySem{M}z_1\dots z_n, \tySem{M}z_1 \dots z_n) \in \rho\
  \tyPrimNm{M}\ (e)$. Thus there exist $x_1,...,x_n$ such that $e =
  e_1^{x_1}\dots e_n^{x_n}$. But we also know that $e =
  i_1^{b_1}...i_m^{b_m}$. By the cancellation property of free
  monoids, we learn that $x_1,...,x_n$ is the solution we need.
\end{proof}

We can further assume that there are other operations that distribute
over the multiplication of the monoid, for example semi-ring
operations (e.g.~relational algebra).

Examples
\begin{enumerate}
\item Finite sets of stuff.
\item Let $\mathcal{M} = \Sigma^* \times
  \mathcal{P}_{\mathit{fin}}(C)$, where $\Sigma$ is some alphabet, and
  $C$ is some set of taint markers. The centre of this monoid is
  $\{([], c) \sepbar c \in \mathcal{P}_{\mathit{fin}}(C) \}$.
\end{enumerate}
